# Cost Methods

## Unit Credit Method
Allocates costs based on benefits accrued each year under the plan formula.

**Normal Cost:**
```
NC = present_value_of_benefit_accrued_in_current_year
```

**Actuarial Liability:**
```
AL = present_value_of_benefits_accrued_to_date
```

## Projected Unit Credit Method
Like Unit Credit but incorporates expected future salary increases.

**For final average pay plans:**
```
Projected_Annual_Benefit = k * (years_of_service) * (1/n) * Σ(final_n_years_salary)
```
Where:

- `k` = benefit multiplier (e.g., 0.015 for 1.5%)
- `n` = averaging period

```{r}
#| label: benefit-calculation
# Final average pay benefit calculation
calculate_final_average_benefit <- function(
  years_of_service,
  final_salaries,
  benefit_multiplier
) {
  average_salary <- mean(final_salaries)
  return(years_of_service * average_salary * benefit_multiplier)
}

# Example
final_5_salaries <- c(80000, 82000, 84000, 86000, 88000)
benefit_example <- calculate_final_average_benefit(30, final_5_salaries, 0.015)
cat("Annual benefit:", benefit_example)
```

## Cash Balance Plans

Cash balance plans are hybrid plans that combine features of defined benefit and defined contribution plans. They maintain individual account balances while providing guaranteed returns and conversion to annuities.

### Cash Balance Account Structure

```{r}
#| label: cb-helper-functions-ch4
# Helper function for survival probability (simplified version)
# Full version is in Chapter 7 - Mortality Tables
calculate_survival_probability <- function(
  mortality_table = NULL,
  age_x,
  years_t
) {
  # Simplified survival probability calculation
  if (is.null(mortality_table)) {
    # Use simplified survival curve if no mortality table
    base_survival <- 0.99  # Annual survival probability at young age
    aging_factor <- 0.995  # Rate of decline
    annual_survival <- base_survival * aging_factor^(age_x - 30)
    return(annual_survival^years_t)
  }
  
  # Check if ages are in the table
  if (!age_x %in% mortality_table$age || 
      !(age_x + years_t) %in% mortality_table$age) {
    # Fallback to simplified calculation
    annual_survival <- 0.99 * 0.995^(age_x - 30)
    return(max(0.01, annual_survival^years_t))
  }
  
  # Calculate cumulative survival
  survival <- 1
  for (t in 0:(years_t - 1)) {
    current_age <- age_x + t
    if (current_age %in% mortality_table$age) {
      idx <- which(mortality_table$age == current_age)
      survival <- survival * mortality_table$p_x[idx]
    } else {
      survival <- survival * 0.98  # Default if age not found
    }
  }
  
  return(survival)
}
```

```{r}
#| label: cash-balance-structure
# Cash balance account tracking
create_cash_balance_account <- function(
  participant_id,
  entry_age,
  entry_salary,
  pay_credit_rate = 0.05,  # 5% of salary
  interest_credit_rate = 0.04  # 4% guaranteed return
) {
  account <- list(
    participant_id = participant_id,
    entry_age = entry_age,
    current_age = entry_age,
    entry_salary = entry_salary,
    current_salary = entry_salary,
    pay_credit_rate = pay_credit_rate,
    interest_credit_rate = interest_credit_rate,
    account_balance = 0,
    transactions = data.frame(
      year = numeric(),
      age = numeric(),
      salary = numeric(),
      pay_credit = numeric(),
      interest_credit = numeric(),
      ending_balance = numeric()
    )
  )
  
  class(account) <- "CashBalanceAccount"
  return(account)
}

# Annual update function for cash balance account
update_cash_balance_account <- function(
  account,
  new_salary = NULL,
  salary_increase_rate = 0.03
) {
  # Update age
  account$current_age <- account$current_age + 1
  
  # Update salary
  if (is.null(new_salary)) {
    account$current_salary <- account$current_salary * (1 + salary_increase_rate)
  } else {
    account$current_salary <- new_salary
  }
  
  # Calculate credits
  pay_credit <- account$current_salary * account$pay_credit_rate
  interest_credit <- account$account_balance * account$interest_credit_rate
  
  # Update balance
  new_balance <- account$account_balance + pay_credit + interest_credit
  
  # Record transaction
  new_transaction <- data.frame(
    year = nrow(account$transactions) + 1,
    age = account$current_age,
    salary = account$current_salary,
    pay_credit = pay_credit,
    interest_credit = interest_credit,
    ending_balance = new_balance
  )
  
  account$transactions <- rbind(account$transactions, new_transaction)
  account$account_balance <- new_balance
  
  return(account)
}

# Example: Create and project cash balance account
cb_account <- create_cash_balance_account(
  participant_id = 1,
  entry_age = 30,
  entry_salary = 50000,
  pay_credit_rate = 0.05,
  interest_credit_rate = 0.04
)

# Project account for 10 years
for (i in 1:10) {
  cb_account <- update_cash_balance_account(cb_account)
}

cat("Cash Balance Account After 10 Years\n")
cat("===================================\n")
print(tail(cb_account$transactions, 5))
cat("\nFinal Account Balance: $", 
    round(cb_account$account_balance, 2), "\n")
```

### Cash Balance Liability Calculations

```{r}
#| label: cash-balance-liability
# Calculate present value of cash balance benefit
calculate_cb_liability <- function(
  current_balance,
  current_age,
  retirement_age,
  pay_credit_rate,
  interest_credit_rate,
  discount_rate,
  salary,
  salary_growth_rate,
  mortality_table = NULL,
  conversion_factor = 12  # Annuity conversion factor
) {
  
  years_to_retirement <- retirement_age - current_age
  
  if (years_to_retirement <= 0) {
    # Already retired - return current balance
    return(current_balance)
  }
  
  # Project account balance to retirement
  projected_balance <- current_balance
  projected_salary <- salary
  
  for (year in 1:years_to_retirement) {
    # Apply interest credit
    projected_balance <- projected_balance * (1 + interest_credit_rate)
    
    # Apply pay credit
    projected_salary <- projected_salary * (1 + salary_growth_rate)
    pay_credit <- projected_salary * pay_credit_rate
    projected_balance <- projected_balance + pay_credit
  }
  
  # Discount back to present value
  pv_projected_balance <- projected_balance / 
    (1 + discount_rate)^years_to_retirement
  
  # Apply mortality discount if table provided
  if (!is.null(mortality_table)) {
    survival_prob <- calculate_survival_probability(
      mortality_table, 
      current_age, 
      years_to_retirement
    )
    pv_projected_balance <- pv_projected_balance * survival_prob
  }
  
  return(pv_projected_balance)
}

# Calculate normal cost for cash balance plan
calculate_cb_normal_cost <- function(
  salary,
  pay_credit_rate,
  interest_credit_rate,
  discount_rate,
  current_age,
  retirement_age,
  mortality_table = NULL
) {
  
  # Pay credit for current year
  pay_credit <- salary * pay_credit_rate
  
  # Years to retirement
  years_to_retirement <- retirement_age - current_age
  
  if (years_to_retirement <= 0) {
    return(pay_credit)  # Immediate payout
  }
  
  # Future value of this year's pay credit at retirement
  fv_pay_credit <- pay_credit * 
    (1 + interest_credit_rate)^years_to_retirement
  
  # Present value using valuation discount rate
  pv_pay_credit <- fv_pay_credit / 
    (1 + discount_rate)^years_to_retirement
  
  # Apply mortality if provided
  if (!is.null(mortality_table)) {
    survival_prob <- calculate_survival_probability(
      mortality_table, 
      current_age, 
      years_to_retirement
    )
    pv_pay_credit <- pv_pay_credit * survival_prob
  }
  
  return(pv_pay_credit)
}

# Example calculations
cb_liability <- calculate_cb_liability(
  current_balance = 100000,
  current_age = 45,
  retirement_age = 65,
  pay_credit_rate = 0.05,
  interest_credit_rate = 0.04,
  discount_rate = 0.06,
  salary = 80000,
  salary_growth_rate = 0.03
)

cb_normal_cost <- calculate_cb_normal_cost(
  salary = 80000,
  pay_credit_rate = 0.05,
  interest_credit_rate = 0.04,
  discount_rate = 0.06,
  current_age = 45,
  retirement_age = 65
)

cat("\nCash Balance Plan Valuation\n")
cat("============================\n")
cat("Current Balance: $100,000\n")
cat("Current Age: 45\n")
cat("PV of Projected Benefit (Liability): $", 
    round(cb_liability, 2), "\n")
cat("Normal Cost: $", round(cb_normal_cost, 2), "\n")
```

### Cash Balance vs Traditional DB Comparison

```{r}
#| label: cb-db-comparison
# Compare cash balance to traditional DB for same participant
compare_cb_to_db <- function(
  entry_age = 30,
  current_age = 45,
  retirement_age = 65,
  entry_salary = 50000,
  current_salary = 80000,
  salary_growth_rate = 0.03,
  
  # Cash Balance parameters
  cb_pay_credit_rate = 0.05,
  cb_interest_credit_rate = 0.04,
  
  # Traditional DB parameters  
  db_benefit_multiplier = 0.015,
  
  # Valuation parameters
  discount_rate = 0.06
) {
  
  # Calculate years of service
  current_service <- current_age - entry_age
  total_service <- retirement_age - entry_age
  
  # Project final salary
  years_to_retirement <- retirement_age - current_age
  final_salary <- current_salary * 
    (1 + salary_growth_rate)^years_to_retirement
  
  # Traditional DB benefit and liability
  db_annual_benefit <- total_service * final_salary * db_benefit_multiplier
  db_accrued_benefit <- current_service * final_salary * db_benefit_multiplier
  
  # Assume annuity factor of 12 for simplicity
  annuity_factor <- 12
  db_liability <- db_accrued_benefit * annuity_factor / 
    (1 + discount_rate)^years_to_retirement
  
  # Cash Balance projection
  # Simulate account growth from entry to current
  cb_balance <- 0
  sim_salary <- entry_salary
  
  for (age in entry_age:(current_age - 1)) {
    # Pay credit
    pay_credit <- sim_salary * cb_pay_credit_rate
    # Interest credit
    interest_credit <- cb_balance * cb_interest_credit_rate
    # Update balance
    cb_balance <- cb_balance + pay_credit + interest_credit
    # Update salary
    sim_salary <- sim_salary * (1 + salary_growth_rate)
  }
  
  # Calculate CB liability
  cb_liability <- calculate_cb_liability(
    current_balance = cb_balance,
    current_age = current_age,
    retirement_age = retirement_age,
    pay_credit_rate = cb_pay_credit_rate,
    interest_credit_rate = cb_interest_credit_rate,
    discount_rate = discount_rate,
    salary = current_salary,
    salary_growth_rate = salary_growth_rate
  )
  
  # Project CB balance to retirement for benefit comparison
  projected_cb_balance <- cb_balance
  projected_salary <- current_salary
  
  for (year in 1:years_to_retirement) {
    projected_cb_balance <- projected_cb_balance * 
      (1 + cb_interest_credit_rate)
    projected_salary <- projected_salary * (1 + salary_growth_rate)
    projected_cb_balance <- projected_cb_balance + 
      projected_salary * cb_pay_credit_rate
  }
  
  cb_annual_benefit <- projected_cb_balance / annuity_factor
  
  # Create comparison
  comparison <- data.frame(
    Plan_Type = c("Traditional DB", "Cash Balance"),
    Current_Liability = c(db_liability, cb_liability),
    Projected_Annual_Benefit = c(db_annual_benefit, cb_annual_benefit),
    Lump_Sum_at_Retirement = c(
      db_annual_benefit * annuity_factor,
      projected_cb_balance
    )
  )
  
  return(list(
    comparison = comparison,
    cb_current_balance = cb_balance,
    details = list(
      entry_age = entry_age,
      current_age = current_age,
      retirement_age = retirement_age,
      current_salary = current_salary,
      years_of_service = current_service
    )
  ))
}

# Run comparison
comparison_results <- compare_cb_to_db()

cat("\nPlan Design Comparison\n")
cat("======================\n")
cat("Participant: Age", comparison_results$details$current_age, 
    "with", comparison_results$details$years_of_service, 
    "years of service\n")
cat("Current Salary: $", comparison_results$details$current_salary, "\n")
cat("Current CB Balance: $", 
    round(comparison_results$cb_current_balance, 2), "\n\n")

# Round only numeric columns for display
comparison_display <- comparison_results$comparison
comparison_display[, -1] <- round(comparison_display[, -1], 2)
print(comparison_display)
```